Lattices of acyclic pipe dreams

### 2023-05-10 acyclicFacets/draft/latticePipeDreams.tex
I added proof in subsection 4.5

\subsection{The acyclic property for $\nu$-Tamari lattices}
\label{subsec:nuTamari}

We close this section by stating an open problem that relates our results to the $\nu$-Tamari lattices introduced by L.-F.~Préville-Ratelle and X.~Viennot in~\cite{PrevilleRatelleViennot}. These posets are indexed by a lattice path $\nu$ consisting of a finite number of north~(N) and east~(E) steps, and coincide with the classical Tamari lattices when $\nu=(NE)^n$.

In~\cite{CeballosPadrolSarmiento}, it was shown that the $\nu$-Tamari lattice can be obtained as the increasing flip poset on pipe dreams~$\pipeDreams(\omega_\nu)$ for an explicit permutation $\omega_\nu$ associated to $\nu$. The permutations of the form~$\omega_\nu$ can be easily characterized as follows. We refer to~\cite{CeballosPadrolSarmiento} for details.

The \defn{Rothe diagram} of a permutation~$\omega$ is the set $\set{(\omega(j),i)}{i<j,\ \omega(i)>\omega(j)}$ in matrix notation.
A permutation $\omega$ is called \defn{dominant} if its Rothe diagram is the Ferrer's diagram of a partition located at the top left corner $(1,1)$. Equivalently, a permutation~$\omega$ is dominant if and only if it is 132-avoiding. 
For a permutation~$\omega = \omega_1 \dots \omega_n \in \fS_n$, we denote by~$0\omega$ the permutation~$0 \omega_1 \dots \omega_n$ of~$\{0, 1, \dots, n\}$, and we consider here pipe dreams with pipes indexed by~$\{0, 1, \dots, n\}$.
%For instance, for the reversed permutation~$\omega = n \dots 2 1$, the permutation~$0\omega$ is the permutation~$\rho_n$ considered in \cref{exm:Tamari1}.
%The permutations $\omega_\nu$ associated to lattice paths $\nu$ are those of the form~$1\oplus w$ for some dominant permutation $w$, see~\cite{CeballosPadrolSarmiento}.

\begin{openproblem}%[The acyclic property for $\nu$-Tamari lattices]
\label{prob:nuAcyclicProperty}
Show that if $\omega$ is a dominant permutation then all pipe dreams with permutation $0 \omega$ are acyclic. %In particular, the corresponding $\nu$-Tamari lattice is the increasing flip poset on acyclic pipe dreams~$\acyclicPipeDreams(1\oplus \omega)$.
\end{openproblem}

From \cref{exm:Tamari1}, we know that this property holds when $\omega=n\dots21$ is the reverse permutation. 
In this case, the increasing flip poset on~$\pipeDreams(0 \omega)$ is the classical Tamari lattice. Removing vertex $0$ from the contact graph of a pipe dream~$ P\in \pipeDreams(0 \omega)$ returns the binary tree corresponding to $P$ (with the leaves removed and the internal nodes labeled from $1$ to $n$).
The edges of the binary tree are oriented going away from the root of the tree.  
See \cref{fig:bijection}.
We have also checked this property computationally for all dominant permutations up to $n=8$.

Applying our Theorem~\ref{thm:pipeDreamQuotient}, we get the following consequence.

\begin{proposition}
If the statement of Open Problem~\ref{prob:nuAcyclicProperty} holds, then the $\nu$-Tamari lattice is a lattice quotient of the interval~$[e,\omega_\nu]$.
\end{proposition}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\input{nuTamari}

%
%\subsection{The acyclic property for $\nu$-Tamari lattices}
%\label{subsec:nuTamari}
%
%We close this section by stating an open problem that relates our results to the $\nu$-Tamari lattices introduced by L.-F.~Préville-Ratelle and X.~Viennot in~\cite{PrevilleRatelleViennot}. These posets are indexed by a lattice path $\nu$ consisting of a finite number of north~(N) and east~(E) steps, and coincide with the classical Tamari lattices when $\nu=(NE)^n$.
%
%In~\cite{CeballosPadrolSarmiento}, it was shown that the $\nu$-Tamari lattice can be obtained as the increasing flip poset on pipe dreams~$\pipeDreams(\omega_\nu)$ for an explicit permutation $\omega_\nu$ associated to $\nu$. The permutations of the form~$\omega_\nu$ can be easily characterized as follows. We refer to~\cite{CeballosPadrolSarmiento} for details. 
%
%The \defn{Rothe diagram} of a permutation~$\omega$ is the set $\set{(\omega(j),i)}{i<j,\ \omega(i)>\omega(j)}$ in matrix notation.
%A permutation $\omega$ is called \defn{dominant} if its Rothe diagram is the Ferrer's diagram of a partition located at the top left corner $(1,1)$. Equivalently, a permutation~$\omega$ is dominant if and only if it is 132-avoiding. 
%For a permutation~$\omega = \omega_1 \dots \omega_n \in \fS_n$, we denote by~$0\omega$ the permutation~$0 \omega_1 \dots \omega_n$ of~$\{0, 1, \dots, n\}$, and we consider here pipe dreams with pipes indexed by~$\{0, 1, \dots, n\}$.
%%For instance, for the reversed permutation~$\omega = n \dots 2 1$, the permutation~$0\omega$ is the permutation~$\rho_n$ considered in \cref{exm:Tamari1}.
%%The permutations $\omega_\nu$ associated to lattice paths $\nu$ are those of the form~$1\oplus w$ for some dominant permutation $w$, see~\cite{CeballosPadrolSarmiento}. 
%
%\begin{openproblem}%[The acyclic property for $\nu$-Tamari lattices]
%\label{prob:nuAcyclicProperty}
%Show that if $\omega$ is a dominant permutation then all pipe dreams with permutation $0 \omega$ are acyclic. %In particular, the corresponding $\nu$-Tamari lattice is the increasing flip poset on acyclic pipe dreams~$\acyclicPipeDreams(1\oplus \omega)$.
%\end{openproblem}
%
%From \cref{exm:Tamari1}, we know that this property holds when $\omega=n\dots21$ is the reverse permutation. 
%In this case, the increasing flip poset on~$\pipeDreams(0 \omega)$ is the classical Tamari lattice. Removing vertex $0$ from the contact graph of a pipe dream~$ P\in \pipeDreams(0 \omega)$ returns the binary tree corresponding to $P$ (with the leaves removed and the internal nodes labeled from $1$ to $n$).
%The edges of the binary tree are oriented going away from the root of the tree.  
%See \cref{fig:bijection}.
%We have also checked this property computationally for all dominant permutations up to $n=8$. 
%
%Applying our Theorem~\ref{thm:pipeDreamQuotient}, we get the following consequence.
%
%\begin{proposition}
%If the statement of Open Problem~\ref{prob:nuAcyclicProperty} holds, then the $\nu$-Tamari lattice is a lattice quotient of the interval~$[e,\omega_\nu]$.
%\end{proposition}
